---
title: 
date: "2025-02-03"
output:
  html_document:
   toc: true
   toc_float: true
---

# Getting Started with R - Part 2

## Step 1: Understanding RStudio Interface
RStudio consists of four main panes:

1. **Source Pane** - This is where you write and edit scripts.
2. **Console Pane** - Displays executed commands and their outputs. The Source Pane allows you to write and edit scripts that can be saved and reused, while the Console Pane executes code immediately but does not save it, meaning any code written directly in the console will be lost once the session ends.
3. **Environment/History Pane** - Shows all the objects in your work space and your command history. In R, objects can be many things, such as vectors, matrices, data sets, functions.You can manually remove all objects at once by clicking the "broom" icon at the top of the pane.
4. **Files/Plots/Packages/Help Pane** - Used for managing files, viewing plots, handling packages, and accessing help documentation. The key tabs to focus on are Files, Plots, and Help. The Files tab allows you to browse your computer's folders and set the working directory by selecting "More" -> Set as Working Directory, although it's generally more efficient to use the command setwd(). In the Plots tab, you can view your generated plots and clear them by clicking the "broom" icon at the top of the pane. The Help tab provides documentation for R functions, datasets, and other objects. To access these, simply type ? before the name of the function, dataset, or object.

## Step 2: Customized Settings 

Navigate to Tools > Global Options to personalize your workspace. Here, you can adjust panel layouts, command themes, and more.

## Step 3: Built-in Datasets

R includes several built-in datasets that are ideal for practice and demonstration purposes. To view the full list, use the command data(). You can also load individual datasets into the environment by using data("Name_of_the_dataset").

```{r}
data() # opens a new window in the Source Pane

```
```{r}
data(cars)
data("iris")
data("AirPassengers")
```

## Step 4: Managing Objects in the R Environment

You can view all objects currently in your R environment using the ls() command. To remove a specific object (e.g., a dataset), use the command rm(Name_of_the_object). If you'd like to clear the entire environment, use rm(list = ls()).

```{r}
ls()  # List all objects in the environment

rm(cars)  # Remove a specific object (e.g., 'cars' dataset)
rm(list = ls())  # Remove all objects from the environment
```

## Step 5: Important Keyboard Shortcuts
Here are some useful keyboard shortcuts for RStudio:
  
### Windows
- **Run Selected Code**: `Ctrl + Enter`
- **Comment/Uncomment Lines**: `Ctrl + Shift + C`
- **Insert Code Section**: `Ctrl + Shift + R`
- **Open Command Palette**: `Ctrl + Shift + P`
- **Assignment operator**: `Alt + -`

### macOS
- **Run Selected Code**: `Cmd + Enter`
- **Comment/Uncomment Lines**: `Cmd + Shift + C`
- **Insert Code Section**: `Cmd + Shift + R`
- **Open Command Palette**: `Cmd + Shift + P`
- **Assignment operator**: `Option + -`

You can customize these shortcuts by going to Tools -> Modify Keyboard Shortcuts.

## Step 6:Installing and Loading Packages

R has many built-in functions, but often you’ll need additional packages for specialized tasks.  Fortunately, the R community has developed thousands of packages to make your work easier, and many of them are available through CRAN (the Comprehensive R Archive Network) or GitHub.

To install a package, you only need to do it once (e.g., after reinstalling R). You can install a package by typing install.packages("package_name") in the console or script. Alternatively, go to Tools -> Install Packages, enter the package name, and click Install. Be sure to check Install dependencies so RStudio automatically installs any required packages.

Once installed, you need to load the package each time you want to use it. You can do this using either the library() or require() functions. (library() stops execution if a package is missing, while require() returns FALSE but allows the code to continue.)

```{r}
# install.packages("tidyverse")
# library(tidyverse)
```


