---
title: 
date: "2025-02-03"
output:
  html_document:
   toc: true
   toc_float: true
---
# Getting Started with R - Part 3
```{r}
knitr::opts_chunk$set(results = "hide")
```
## Step 1: Hello World!
To output text in R, enclose it in single or double quotes. For numbers, simply type them:
```{r}
"Hello World!"
'Hello World!'
print("Hello World")
25
print(25)
```
R can also be used as a calculator:

```{r}
2+3
8^2
```
## Step 2: Creating objects
R is an object-oriented programming language, meaning everything you create or manipulate can be stored in an object and accessed later. Whether you’re loading a dataset, modifying data, running analyses, creating plots, or even writing functions, you store everything in objects. This is a key difference from software like Stata or SPSS, where you work with a single active dataset.

An object stores a value or data structure, which can be accessed later using its name. To create an object, assign a value using the "<-" operator . Viewing the object in the console returns its stored value.

```{r}
name1 <- "Bram"
name2 <- "Sara"
age <- 35
1980 -> birthyear
name1
age

name2 <- name1
name <- name2
name

birthyear1 <- birthyear2 <- 2002
```

The = operator can be used as an alternative to <- for assignment in R, but it is generally discouraged for beginners. R prioritizes <- over = in assignment operations.

```{r}
#age1 <- age2 = 25
```

You can combine two or more elements using the paste() function.
```{r}
text <- "fun"
paste("this class is", text)

text1 <- "this class is"
text2 <- "fun"
paste(text1, text2)

age <- 36
paste("I am", age, "years old.")
myage <- paste("I am", age, "years old.")
```


## Step 3: Data types

- Numerics: Decimal values like 3.5.
- Integers: Whole numbers like 3, which are also a type of numeric.
- Logical: Boolean values, TRUE or FALSE.
- Characters: Text or string values.
- Factors: Cateogrical data (useful for grouping, plotting, and statistical models)
- Date & Time: Date stores only the date, while POSIXct/POSIXlt includes both date and time.
```{r}
num_value <- 3.5 
num_value
class(num_value)  

int_value <- 4L  
int_value
class(int_value)  

bool_value <- TRUE  
bool_value <- F
bool_value

char_value <- "Hello, R!"  
char_value

marital_status <- factor(c("Single", "Married", "Divorced"))
marital_status
#If you don’t specify levels, R automatically alphabetically sorts the unique values.
# Factors are stored as integers internally
as.integer(marital_status)

date_value <- as.Date("2025-02-06")  
date_value

datetime_value <- as.POSIXct("2025-02-06 14:30:00")  
datetime_value
```

You can inspect the data type of an object using the class() function:
```{r}
class(age)
class(int_value)
class(bool_value)
class(datetime_value)
```

If you want to check whether an object is of a specific type, you can use these functions:
```{r}
is.numeric(bool_value)
is.integer(bool_value)
is.character(bool_value)
is.logical(bool_value)
```

## Step 4: Data structures 

There are several fundamental data structures used for storing and organizing data. Here’s a brief explanation of each:

### 1. Vectors
In R, vectors are one-dimensional sequence of elements of the same type (e.g., numeric, character, logical), and it is treated as a single row or column of data. They are most basic data structure in R. They can be created using the function c().

```{r}
# Creating a numeric vector
integer_vector <- c(10, 20, 30, 40, 50)
integer_vector <- c(10:20)
integer_vector <- 1:5
integer_vector #or print(integer_vector)

```

You can access elements using indexing [].

```{r}
# Accessing elements using indexing 
integer_vector[3] 

## Access the first and third item 
integer_vector[c(1,3)] 

## Access the first three items
integer_vector[1:3]

# You can also use negative index numbers to retrieve all elements except for the ones specified.
integer_vector[-1]

```

You can change the value of a specific element, by referring to the index number:
```{r}
# Change 20 to 25
integer_vector[2] <- 25
integer_vector
```


```{r}
# Create a vector containing both numbers and strings
string_vector <- c(1, "Sanne", "Jan", 45)

string_vector #or print(string_vector)

# Note that because the vector contains strings, the numbers are also treated as strings
```

### 2. Matrices
Matrices are two-dimensional arrays with rows and columns. All elements in a matrix must be of the same data type. Matrices can be created using the function matrix().

```{r}
my_matrix <- matrix(c(1, 5, 7, 9, 15, 36), nrow = 2, ncol = 3)
print(my_matrix)
```

You can access elements by row and column index [row, column].
```{r}
my_matrix[1,3]
```

### 3. Data Frames
Data frames are two-dimensional data structures that resemble tables or spreadsheets.They can store different types of data in different columns.You can check the data types of columns of the data frame using str() (base R) or  glimpse() (dplyr).

Data frames are created using the function data.frame().

```{r}
id <- c(1, 2, 3, 4, 5)
name <- c("Sanne", "Jan", "Mila", "Bram", "Thijs")
age <- c(25, 30, 35, 40, 45)
score <- c(85, 90, 88, 94, 80)

my_df <- data.frame(id, name, age, score)
print(my_df)

new_df <- data.frame(age, eyecolor <- c("blue", "green", "brown", "blue", "brown"))
print(my_df)
```


```{r}
str(my_df)
dplyr::glimpse(my_df)
```

You can access rows and columns by name or by index: 
```{r}
my_df$age
# OR
my_df[["age"]]
#OR
my_df[,"age"]
#OR
my_df[[3]]
#OR 
my_df[,3]


my_df[2, 3]
#OR
my_df[2, "age"]
```

### 4. Lists
Lists are ordered collections of objects, which can be of different types and structures. Each element of a list can be a vector, matrix, data frame, tibble, or even another list.
Lists can be created using the list() function.
```{r}
my_list <- list(
  numeric_vector = c(4, 7, 13),
  character_vector = c("ab", "bb", "cb"),
  matrix_data = matrix(1:4, nrow = 2, ncol = 2),
  dataframe = data.frame(x = 1:3, y = c("a", "b", "c"))
)

print(my_list)
```

You can access elements by index or name:
```{r}
my_list[[1]] 
#OR
my_list[["numeric_vector"]] 
#OR
my_list$numeric_vector
```

## Step 5: Some key packages in R 

### 1. Base R  
Base R comes pre-installed and provides essential functions for **data processing, statistical analysis, and basic plotting**. It builds a strong foundation in R by encouraging users to understand core concepts and solve problems with a general toolkit before relying on specialized packages.  

**Common functions:** `mean()`, `sum()`, `aggregate()`, `lm()` (linear modeling), `plot()`, and more.  

### 2. Tidyverse  
Tidyverse is a collection of R packages designed for **data manipulation and visualization**, following a consistent and intuitive approach.  

- **ggplot2** – Data visualization  
- **dplyr** – Data manipulation (`filter()`, `select()`, `mutate()`, `summarize()`)  
- **tidyr** – Data tidying (reshaping, pivoting)  

- **tibble** – Modern data frames  

### 3. data.table  
`data.table` is an optimized package for **fast and efficient data manipulation**, particularly suited for **big data**. Filtering, grouping, and summarizing operations run significantly faster compared to **Base R** and **dplyr**.  

---

### Example: Filtering Rows (Age > 30), Selecting Columns, and Sorting (Syntax Examples)  

#### Base R
```{r}

# Filter rows where age > 30
filtered_df <- my_df[my_df$age > 30, ]  

# Select columns (name and age)
selected_df <- filtered_df[, c("name", "age")]

# Sort by age
sorted_df <- selected_df[order(selected_df$age), ]

# Display result
sorted_df
```

#### Tidyverse
```{r}
library(dplyr)

my_df %>%
  filter(age > 30) %>%
  select(name, age) %>%
  arrange(age)
```


#### data.table
```{r}
library(data.table)

DT <- as.data.table(my_df)
DT[age > 30, .(name, age)][order(age)]
```

