---
title: "Data and Measurement"
output:
  html_document:
    theme: 
       bootswatch: minty
    df_print: kable
    code_folding: hide 
---

# Tutorial 2
In this exercise tutorial, we will tackle various data wrangling and visualization tasks using the dataset named "oxis2013ukda.dta". The dataset originates from the longest-running academic survey of internet use in Britain, detailing the evolution of internet use from 2003 to the present day. It encompasses nearly 796 variables measuring internet activities, attitudes, and effects. Before proceeding to the exercises, you can review some notes below regarding data types and structures in R.

##  {.tabset .tabset-fade .tabset-pills}
### Overview of data types and structures
#### Data types

- Numeric: Used to represent numeric values, including both integers and floating-point numbers (double).
- Integer: Specifically for storing whole numbers without decimal places.
- Character: Used to represent text data, including strings of characters.
- Logical: Represents Boolean values, either TRUE or FALSE.
- Factor: Represents categorical data with predefined levels or categories.
- Date: Specifically for representing dates without time components (not covered in this tutorial)

#### Missing values

In R, missing values are denoted by 'NA' and are not treated as regular values. They can disrupt any expression you attempt to use them in.

To identify missing values, you can employ the function 'is.na()'. Additionally, when assessing a series of values that may include a missing value, remember to specify 'na.rm = TRUE' (the default is 'FALSE') to ensure accurate computations.

#### Data structures

There are several fundamental data structures used for storing and organizing data. Here's a brief explanation of each:

<b>Vectors</b>:
The most basic data structure in R. They can hold elements of the same data type and can be created using the function c(). 
```{r}
my_vector <- c (1, "Alice", "Bob", 45)
print(my_vector)
```

You can access elements by index [].

```{r}
my_vector[3]

# Access two elements at positions 2 and 4
my_vector[c(2, 4)]
```

<b>Matrices</b>:
Matrices are two-dimensional arrays with rows and columns.All elements in a matrix must be of the same data type. Matrices can be created using the function matrix().

```{r}
my_matrix <- matrix(c(1, 5, 7, 9, 15, 36), nrow = 2, ncol = 3)
print(my_matrix)
```
You can access elements by row and column index [row, column].
```{r}
my_matrix[1,3]
```

<b>Data Frames</b>:
Data frames are two-dimensional data structures that resemble tables or spreadsheets.
They can store different types of data in different columns.You can check the data types of columns of the data frame using str() (base R) or  glimpse() (dplyr).

Data frames are created using the function data.frame().

```{r}
my_df <- data.frame(
  id = c(1, 2, 3),
  name = c("Alice", "Bob", "Charlie"),
  age = c(25, 30, 35)
)
print(my_df)
```
```{r}
str(my_df)
dplyr::glimpse(my_df)
```

You can access columns by name: 
```{r}
my_df$age
# OR
my_df[["age"]]
```

You can access rows and columns by index: 
```{r}
my_df[2, 3]
#OR
my_df[2, "age"]
```

<b>Tibbles</b>:
Tibbles are enhanced data frames for streamlined data processing. Tibbles, belonging to the tidyverse ecosystem, offer a modern approach to handling tabular data.

Advantages of tibbles include:

- Limited default output rows for easier large dataset display.
- Retention of original column names without conversion to character vectors.
- Prevention of automatic string-to-factor conversions.
- Preservation of result class when subdividing tibbles with the $ operator.
- Tibbles can be created using functions like tibble() or as_tibble() from the tibble package.

```{r}
tibble::tibble(
  id = c(1, 2, 3),
  name = c("Alice", "Bob", "Charlie"),
  age = c(25, 30, 35)
)
```
You can access columns by name (same as data frames) and access rows and columns by index (same as data frames).


<b>Lists</b>:
Lists are ordered collections of objects, which can be of different types and structures. Each element of a list can be a vector, matrix, data frame, tibble, or even another list.
Lists can be created using the list() function.
```{r}
my_list <- list(
  numeric_vector = c(4, 7, 13),
  character_vector = c("ab", "bb", "cb"),
  matrix_data = matrix(1:4, nrow = 2, ncol = 2),
  dataframe = data.frame(x = 1:3, y = c("a", "b", "c"))
)

print(my_list)
```

You can access elements by index or name:
```{r}
my_list[[1]] 
#OR
my_list[["numeric_vector"]] 
#OR
my_list$numeric_vector

```


### Exercise 1 - Data wrangling
1) Load the "oxis2013ukda.dta"  into the environment.
Note that the data is saved in stata file format and you can use "haven" package.

2) Check the class of the dataset. What do tbl_df and tbl mean?

3) Check the structure of the following four variables:`trmost`, `trgov`, `accmo`, and `dg4`. Display the variable labels separately for each variable (tip: utilize "attr" function)

4) Assign the label "Q.L1 What is your highest educational or vocational qualification? 4 ctg (Generated variable)" to the variable "dg4".

5) What are the numerical values corresponding to the categories of variable dg4?

6) Convert the variable dg4 to a factor and save it as a new variable named "education". Rename the categories as follows:
"No qual." to "No diploma"
"Secondary" to "Secondary education"
"Further ed" to "Further education college"
"Higher ed." to "Higher education"

7) Merge the "Further education college" and "Higher education" categories of the variable education into one category named "More than secondary".

8) Determine the number of missing values in the four variables using the functions "is.na()" and "sum()".

9) Identify the unique values of `trmost`(You can utilize the "unique()" function or the "table()" function).

10) Compute the mean of `trmost`, `trgov` and `accmo`. Remember to remove NAs while using mean(). Aim to condense the command into one or two lines while obtaining the mean for all three variables (for example use sapply(), or colMeans()).

### Exercise 2 - Data visualization
1) Visualize `trmost`, `trgov`, `accmo`, and `education` separately. Note: When plotting, ggplot2 ignores NA values and only plots the available data. However, when the variable is a factor, ggplot plots NA as a category.

2) Combine the four plots from step 1 (Tip: Utilize cowplot::plot_grid)

3) Visualize the relationship between trgov and accmo. Would a scatterplot work? If not, what other ways are there to visualize this relationship?

4) Visualize the mean internet duration in months for each value of trgov.

5) Recode trmost into two categories (high trust vs low trust) and visualize the relationship between education (3 categories) and this new trmost variable.

6) Using the 1.5 * IQR (Interquartile Range) rule, determine the number of outliers present in the variable 'accmo'. You can calculate this manually. To observe the smallest and largest values, you can utilize the table() function or display the head and tail of the sorted variable using head() and tail() functions.

7) Generate a boxplot for the variable 'accmo' and verify if the number of outliers you identified in step 6 corresponds to the outliers depicted in this plot.


