---
title: "Data and Measurement"
output:
  html_document:
    theme: 
       bootswatch: minty
    df_print: kable
    code_folding: hide 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results = "hide")
```

# Week 1 - Exercises

In the first set of exercises, we will practice creating different data structures in R and using various functions (such as sample() and rnorm()).
In the second set of exercises, we will perform data wrangling tasks using the dataset educ_inc_age_data.csv, which contains information about individuals’ age, income, and education level.

##  {.tabset .tabset-fade .tabset-pills}
### Exercises - Getting started with R  
1. Create a numeric vector with numbers from 1 to 100 called id.
```{r}
# Create a numeric vector called id
id <- 1:100
```

2. Create a numeric vector containing the following numbers: 3, 5, 8, 9, 67.
```{r}
# Create a numeric vector
numeric_vector <- c(3, 5, 8, 9, 67)
```

3. Check whether the vector you have created is actually numeric.
```{r}
is.numeric(numeric_vector)

#OR
class(numeric_vector)
```

4. Create a character vector with apple, banana, orange, pear and persimmon. 
```{r}
# Create a character vector
character_vector <- c("apple", "banana", "orange", "pear", "persimmon")
```

5. Check whether the vector you have created is a character vector.
```{r}
is.character(character_vector)
#OR
class(character_vector)
```

6. Create a matrix with 5 rows and 2 columns. The first column contains the following numbers: 3, 5, 8, 9, 67 and the second column contains the following strings: "apple", "banana", "orange", "pear", "persimmon".
```{r}
my_matrix <- matrix(c(3, 5, 8, 9, 67, "apple", "banana", "orange", "pear", "persimmon"), nrow = 5, ncol = 2)

# Note that matrices can only contain a single class of data (numeric, character etc.), while data frames can consist of many different classes of data. In this matrix, numbers are treated automatically as characters. 
```

7. Create the same matrix as in step 6 by combining the two vectors you created in step 2 and 3. 
```{r}
my_matrix_combined <- matrix(c(numeric_vector, character_vector), nrow = 5, ncol = 2, byrow = FALSE)
```

8. Would it be possible to create a data frame by combining the three vectors you created in steps 1, 2 and 3? Try it out and see what happens. Note that the vector you created in step 1 has a different length than the vectors in steps 2 and 3.
```{r}
# Create a data frame from the three vectors
my_data <- data.frame(id, numeric_vector, character_vector)
# it works but what happens is that the shorter vectors are recycled to match the length of the longer vector. (note that this happens only if the shorter vector's length is a factor of the longest vector’s length.)
```

9. Create a data frame consisting of 50 rows and three variables: `name`, `age`, and `city`.
- The `name` variable should include names randomly selected from the list: "John", "Alice", "Emma", "Janneke", and "Jelle".
- The `age` variable should contain random draws from a normal distribution with a mean of 0 and a standard deviation of 5.
- The `city` variable should be randomly selected from the cities: "New York", "Los Angeles", and "Chicago". The probability of being chosen should be 0.3 for "New York" 0.6 for "Chicago", and 0.1 for "Los Angeles". 
After creating the data frame, check the frequency table of the variable `city`.
```{r}
my_data <- data.frame(
  name = sample(c("John", "Alice", "Emma", "Janneke", "Jelle"), 50, replace = TRUE),
  age = rnorm(50, mean = 0, sd = 5),
  city = sample(c("New York", "Los Angeles", "Chicago"), 50, replace = TRUE, prob = c(0.3, 0.1, 0.6)))

# Print the first few rows of the created data frame
head(my_data)

# Check  the frequency of cities
city_frequency <- table(my_data$city)
city_frequency
```

10. Execute your code from step 9 again and check the frequencies of the variable `city` or the values of the variables `name` and `age` once more. Do you see any differences?
```{r}
my_data <- data.frame(
  name = sample(c("John", "Alice", "Emma", "Janneke", "Jelle"), 50, replace = TRUE),
  age = rnorm(50, mean = 0, sd = 5),
  city = sample(c("New York", "Los Angeles", "Chicago"), 50, replace = TRUE, prob = c(0.3, 0.1, 0.6)))

# Print the first few rows of the created data frame
head(my_data)

# Check  the frequency of cities
city_frequency <- table(my_data$city)
city_frequency
# yes, the frequencies have changed because we used a random sample without setting a seed!!!
```

11. To ensure reproducibility, set the seed and then run your code again.
```{r}
set.seed(123) # When you set the seed to a specific value, you ensure that the same sequence of random numbers is generated every time you run your code. This is particularly useful for debugging and sharing code, as it allows others to reproduce your results exactly.

my_data <- data.frame(
  name = sample(c("John", "Alice", "Emma", "Janneke", "Jelle"), 50, replace = TRUE),
  age = rnorm(50, mean = 0, sd = 5),
  city = sample(c("New York", "Los Angeles", "Chicago"), 50, replace = TRUE, prob = c(0.3, 0.1, 0.6)))

# Print the first few rows of the created data frame
head(my_data)

# Check  the frequency of cities
city_frequency <- table(my_data$city)
city_frequency
```

### Exercises - Data wrangling
1. Load the dataset and display its structure, along with its first and last five observations.
```{r include = TRUE}
# Read the dataset into a data frame
educ_inc_age_data <- read.csv("/Users/msimsek/Desktop/UvA/Teaching/Data_and_Measurement_RMSS/2024/W1/Tutorials/educ_inc_age_data.csv")

# Display the structure of the dataset
str(educ_inc_age_data)

# Display the first five obs
head(educ_inc_age_data, n = 5) # By default, head() returns the first 6 rows, but you can specify the number of rows to display by providing the n argument.

#Display the last five obs
tail(educ_inc_age_data, n = 5) ## By default, tail() returns the last 6 rows, but you can specify the number of rows to display by providing the n argument.
```

2. Create a new dataset by filtering to include only individuals with an `income` greater than $50,000.
```{r include = TRUE}
# Filtering data for income greater than $50,000
high_income_data <- educ_inc_age_data[educ_inc_age_data$income > 50000, ]
```

3. How many observations does this new dataset have? 
```{r include = TRUE}
# Counting the number of observations
num_observations <- nrow(high_income_data)
num_observations
```

4. Sort the dataset based on `age` in ascending order.
```{r}
# Sorting data based on age in ascending order
sorted_age_data <- educ_inc_age_data[order(educ_inc_age_data$age), ]
head(sorted_age_data)
```

5. Group the data by `education level` and calculate the mean `income` for each group.
```{r}
# Grouping data by education level and calculating mean income
mean_income_by_education <- aggregate(income ~ education, data = educ_inc_age_data, FUN = mean) 
#income ~ education specifies that we want to aggregate the 'income' variable by the levels of the 'education' variable. data = educ_inc_age_data specifies the dataset where the variables are located. FUN = mean indicates that we want to calculate the mean income for each education level.

mean_income_by_education
```

6. Create a new variable named "age_group" categorizing individuals into three age groups: "Young" (`age` < 30), "Middle-aged" (30 <= `age` < 65), and "Elderly" (`age` >= 65).
```{r}
# Create a new variable 'age_group'
educ_inc_age_data$age_group <- NA  # Initialize the new variable with NA values

# Assign 'Young' to individuals aged less than 30
educ_inc_age_data$age_group[educ_inc_age_data$age < 30] <- "Young"

# Assign 'Middle-aged' to individuals aged between 30 and 65
educ_inc_age_data$age_group[educ_inc_age_data$age >= 30 & educ_inc_age_data$age < 65] <- "Middle-aged"

# Assign 'Elderly' to individuals aged 50 and above
educ_inc_age_data$age_group[educ_inc_age_data$age >= 65] <- "Elderly"

head(educ_inc_age_data)
tail(educ_inc_age_data)
```


7. Add a new variable called age_squared which is the square of the `age` variable
```{r}
# Add a new variable called age_squared which is the square of the age variable
educ_inc_age_data$age_squared <- educ_inc_age_data$age^2

# Print the first few rows of the updated data frame
head(educ_inc_age_data)
```

8. Remove the variable `age_squared` from the dataset.
```{r}
# Remove the variable age_squared from educ_inc_age_data
educ_inc_age_data <- subset(educ_inc_age_data, select = -age_squared)

#OR

#educ_inc_age_data$age_squared <- NULL
```

9. What is the value of `education` for row 14 and value of `age` for row 3?
```{r}
# display the row 14 of the column city
educ_inc_age_data[14,"education"]
educ_inc_age_data[3, "age"]
```

10. Identify the row numbers of observations where `age` is 25 or less. Then, using these row numbers, extract the corresponding income values. 
```{r}
# Find the row numbers for values in the "age" variable smaller than 25
indexes_smaller_than_25 <- which(educ_inc_age_data$age <= 25)
# Extract the income values for these rows
educ_inc_age_data[indexes_smaller_than_25, "income"]

#OR

educ_inc_age_data$income[educ_inc_age_data$age <= 25]
```

11. Convert the variable `age` to character.
```{r}
educ_inc_age_data$age <- as.character(educ_inc_age_data$age)
is.character(educ_inc_age_data$age)
```

12. Convert the variable `age` to numeric
```{r}
educ_inc_age_data$age <- as.numeric(educ_inc_age_data$age)
is.numeric(educ_inc_age_data$age)
```

13. Check for missing values in the dataset.
```{r}
# Checking for missing values
missing_values <- sum(is.na(educ_inc_age_data))
missing_values
```

