---
title: "Data and Measurement"
output:
  html_document:
    theme: 
       bootswatch: minty
    df_print: kable
    code_folding: hide 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results = "hide")
```

# Week 2 - Exercises - Part 1 
In this tutorial, we will explore several descriptive statistics tasks using the dataset educ_inc_age_data.csv. This dataset contains information on individuals' age, income, and education level.

##  {.tabset .tabset-fade .tabset-pills}

### Exercises - Descriptive statistics
1. Load the dataset and display its structure, along with its first and last five observations.
```{r include = TRUE}
# Read the dataset into a data frame
educ_inc_age_data <- read.csv("/Users/msimsek/Desktop/UvA/Teaching/Data_and_Measurement_RMSS/2024/W1/Tutorials/educ_inc_age_data.csv")

# Display the structure of the dataset
str(educ_inc_age_data)

# Display the first five obs
head(educ_inc_age_data, n = 5) # By default, head() returns the first 6 rows, but you can specify the number of rows to display by providing the n argument.

#Display the last five obs
tail(educ_inc_age_data, n = 5) ## By default, tail() returns the last 6 rows, but you can specify the number of rows to display by providing the n argument.
```


2. Compute the mean, median, and mode for the `age` variable.
```{r}
# Mean
mean_age <- mean(educ_inc_age_data$age)
mean_age
# Median
median_age <- median(educ_inc_age_data$age)
median_age

# Mode
# Mode function is not readily available in base R. You can however create our own function (though maybe this is a bit advanced, so we try it later in the course), or use a function from a package, such as DescTool.

mode_age <- DescTools::Mode(educ_inc_age_data$age)
mode_age #the first value we get is the mode and the second one from attr(, "freq") is the frequency of the mode value
```


3. Compute the range for `income` variable.
```{r}
# Range
income_range <- max(educ_inc_age_data$income) - min(educ_inc_age_data$income)
income_range
```


4. Compute the variance and standard deviation for the `age` variable.
```{r}
# Variance
var_age <- var(educ_inc_age_data$age)
var_age

# Standard Deviation
sd_age <- sd(educ_inc_age_data$age)
sd_age
```

5. Compute the 25th, 50th (median), and 75th percentiles, as well as the interquartile range for the `income` variable.
```{r}
# Percentiles
percentiles_income <- quantile(educ_inc_age_data$income, c(0.25, 0.5, 0.75))
percentiles_income

# Interquartile Range
iqr_income <- IQR(educ_inc_age_data$income)
iqr_income
```

6. Compute the Z-scores for the `age` variable.
```{r}
# Z-scores for age
z_scores_age <- (educ_inc_age_data$age - mean_age) / sd_age
z_scores_age

```

7. Compute the Z-score for `income` = 14853
```{r}
# Z-score for income
z_score_income <- (14853 - mean(educ_inc_age_data$income)) / sd(educ_inc_age_data$income)
```

8. Ask for summary statistics for each variable in the dataset using one function. 
```{r}
summary(educ_inc_age_data)

#Later in the course, we will use the psych package for reliability and validity analyses. This package also includes a useful summary function called describe.
psych::describe(educ_inc_age_data)
```


